Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2024-07-31 report/emse24.tex
feat(report): emse24 abstract updates

% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{wrapfig}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent#1
\end{framed}}

@@

\author{Arumoy Shome\and
Lu\`{i}s Cruz\and
  Diomidis Spinellis\and
  Arie van Deursen
}

@@

\begin{abstract}
  The machine learning development lifecycle involves iterative and exploratory processes that heavily rely on feedback mechanisms to ensure data and model integrity. This study investigates the use of implicit and explicit feedback within Jupyter notebooks. We analyse 297.8 thousand Jupyter notebooks from GitHub and Kaggle, identifying key feedback mechanisms such as print statements, outputs from last cell statements, and assertions. Our findings reveal extensive use of implicit feedback for critical design decisions and highlight the limited adoption of explicit feedback mechanisms like assertions. By conducting detailed case studies on selected feedback instances, we uncover the reliance on manual validation and the potential for automated validation to enhance robustness and maintainability. This study underscores the need for improved documentation and testing practices within the ML community, advocating for the integration of feedback mechanisms into the ML development workflow to mitigate technical debt and enhance reproducibility.
\end{abstract}

\keywords{Assertions, Feedback Mechanisms, Jupyter notebooks, SE4AI}

@@

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modelling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

@@

As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\cite{kery2018story,rule2018exploration}. In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones. This transition reduces the reliance on manual verification, thereby enhancing the robustness and maintainability of ML pipelines.

@@

    We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements. We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks. The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence~\cite{replication}.

@@

% The above all motivates the research: but then the intro should also ¡®sell¡¯ the outcomes of the research. These are
% - Identification of over 20 feedback mechanisms, half of them implicit, half of them explicit (your taxonomy)
% - Insights in actual use of these mechanisms
% - Recommendations for better support of these mechanisms in notebook environments / life cycle

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{nb.pdf}
  \caption{Example Jupyter notebook with code and markdown cells, adapted ~\cite[Figure~1]{pimentel2019large-scale}.}
  \label{fig:nb}
\end{figure}

@@

The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)~\cite{pythonast}. The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree and compile it back into executable code, providing a powerful tool for dynamic code manipulation. In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. Print statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

@@

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}. The libraries are selected based on the import analysis conducted by \emph{Psallidas et al.}~\cite{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

@@

% TODO: add citation to online appendix

The results of the study are presented in this section. For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. This is done only to structure the paper in a meaningful manner and for ease of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples. These can be viewed on our online appendix~\cite{appendix}.

@@

% TODO: add the findings from this part of the analysis

@@

\subsubsection{Data shape check ($N = 26$)}
% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)

@@

\subsubsection{Data validation check ($N = 14$)}
% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)

@@

\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)

@@

\subsubsection{Existence check ($N = 8$)}
% NOTE: this is a combination of missing-value-check and existence-check
% (A23, A42, A43, A50, A51, A63, A79, A86)

@@

\subsubsection{Resource check ($N = 7$)}~\label{sec:result-rq2-resource-check}
% (A10, A14, A18, A37, A60, A67 A74)

@@

\subsubsection{Type check ($N = 5$)}
% (A2, A35, A40, A81, A88)

@@

\subsubsection{Mathematical property check ($N = 4$)}
% (A3, A25, A56, A64)

@@

\subsubsection{Batch size check ($N = 3$)}
% (A21, A28, A70)

@@

\subsubsection{Network architecture check ($N = 3$)}
% (A11, A62, A75)

@@

\subsubsection{Data leakage check ($N = 1$)}
% (A33)

@@

\subsubsection{Model performance check ($N = 33$)}
% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)

@@

\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
% (O2, O4, O9, O14, O20, O25, O48)

@@

\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
% (P71, P86, O64, O66, P68, P82, P107)

@@

\subsubsection{Spot check ($N = 5$)}
% (O56, O60, P64, P67, P114)

@@

\subsubsection{Model training check ($N = 4$)}
% (O8, O31, O42, P77)

@@

\subsubsection{Missing value check ($N = 3$)}
% (O12, O36, P74)

Checking for missing values is a critical step in the preprocessing phase of an ML project because it significantly influences the quality and performance of the model. Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}. Many ML algorithms demand complete numerical datasets to perform calculations such as matrix multiplication. Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely. These checks can be performed in code (\emph{P74, L36}) and using visualizations such as a heatmap (\emph{L12}).

@@

\subsubsection{Shape check ($N = 3$)}
% (P4, P32, P117)

@@

\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
% (O6, O10)

@@

Linear ML models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}. Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships among features in the data. In addition to the code, the raw output of the cell is also shown since both case-studies produce a visualization.  In \emph{L6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}. \emph{L10} is a visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.

@@

\subsubsection{Type check ($N = 2$)}
% (O71, P43)

@@

\subsubsection{Execution time check ($N = 1$)}
% (P66)
\begin{lstlisting}[caption={Print statement \emph{P66} used to check the total execution time of training an ML model with various hyper-parameter configurations.}, label={lst:P66}]
print('Total Run Time:')
\end{lstlisting}

While training multiple ML models for a single task is common, choosing the best model considers not only its performance but also its training speed. Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}. Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyperparameter configurations using a grid-search.

@@

\subsubsection{Network architecture check ($N = 1$)}
% (P92)

@@

Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}. Once the code is finalized, it is typically transferred into a script that is run in a production environment. Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration. To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}. In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.

We believe that such structured information can be derived from the feedback mechanisms available in Jupyter notebooks. Our results show that practitioners use assertions, print statements, output of code cells and visualizations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and check for hardware and software dependencies. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

Our results from RQ3 reveal that implicit feedback from print and last cell statements are extensively used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle. Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.

We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\cite{heer2010tour}. By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\cite{yang2021subtle}.

@@

Catching data errors are critical since ML models trained using dirty data will lead to poor and incorrect predictions. This is a cause for concern since these models may be deployed in production and their output may be used by other services which will then also perform poorly. The output of the ML models are often also used to capture new training data, thus leading to a feedback loop which over time will degrade the model performance~\cite{sculley2015hidden,breck2019data}.

\emph{Breck et al. (2017)}~\cite{breck2017ml} view ML training as analogous to compilation and recommend that training data be tested, similar to code in traditional software. They propose a rubric to assess the production readiness of an ML pipeline, which includes several recommendations for data testing. In subsequent work, \emph{Breck et al. (2019)}~\cite{breck2019data} introduce \emph{TensorFlow Data Validation}, an automated tool for data validation that works on principles borrowed from the database community. Data expectations are specified in a schema, and each new data batch is validated against this schema to ensure it meets the expected criteria or prompts schema updates if requirements change.

The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged. This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\cite{wang2020assessing}.

>>>>

% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{natbib}
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{wrapfig}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent#1
\end{framed}}

@@

\author{Arumoy Shome\and
Lu{\'\i}s Cruz\and
  Diomidis Spinellis\and
  Arie van Deursen
}

@@

\begin{abstract}
  The machine learning development lifecycle is characterized by iterative and exploratory processes that rely on feedback mechanisms to ensure data and model integrity. Despite the critical role of feedback in machine learning engineering, no prior research has been conducted to identify and understand these mechanisms. To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyse 2.3 million code cells. We identify three key feedback mechanisms---assertions, print statements and last cell statements---and further categorize them into implicit and explicit forms of feedback. Our findings reveal extensive use of implicit feedback for critical design decisions and the relatively limited adoption of explicit feedback mechanisms. By conducting detailed case studies with selected feedback instances, we uncover the potential for automated validation of critical assumptions in ML workflows using assertions. Finally, this study underscores the need for improved documentation, and provides practical recommendations on how existing feedback mechanisms in the ML development workflow can be effectively used to mitigate technical debt and enhance reproducibility.
\end{abstract}

\keywords{SE4AI, Feedback Mechanisms, Jupyter notebooks, Assertions, Technical Debt}

@@

The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.

The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\citep{betz2018managing}. Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, to data preparation, modelling, evaluation and finally deployment and monitoring~\citep{martinez-plumed2021crisp-dm}. Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.

Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\citep{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}. Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\citep{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.

@@

As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\citep{kery2018story,rule2018exploration}. In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones. This transition reduces the reliance on manual verification, thereby enhancing the robustness and maintainability of ML pipelines.

@@

    We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.

    We create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements. We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks. The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence\footnote{https://doi.org/10.6084/m9.figshare.26372140.v1}.

@@

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\citep{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{nb.pdf}
  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\citep[Figure~1]{pimentel2019large-scale}.}
  \label{fig:nb}
\end{figure}

@@

The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)\footnote{https://docs.python.org/3/library/ast.html}. The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree and compile it back into executable code, providing a powerful tool for dynamic code manipulation. In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class. The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails. Print statements are treated as a call to the built-in \lstinline{print} function. Function calls are represented in the AST by the \lstinline{Call} class. The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.

@@

Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}. We used the pre-existing dataset KGTorrent~\citep{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub. To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks. Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell. Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}. The libraries are selected based on the import analysis conducted by \citet{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub. We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

@@

The results of the study are presented in this section. For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other. This is done only to structure the paper in a meaningful manner and for ease of reporting the results. Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}. Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples. These can be viewed on our online appendix\footnote{https://arumoy.me/shome2023notebook}.

@@

@@

\subsubsection{Data shape check ($N = 26$)}

@@

\subsubsection{Data validation check ($N = 14$)}

@@

\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}

@@

\subsubsection{Existence check ($N = 8$)}

@@

\subsubsection{Resource check ($N = 7$)}~\label{sec:result-rq2-resource-check}

@@

\subsubsection{Type check ($N = 5$)}

@@

\subsubsection{Mathematical property check ($N = 4$)}

@@

\subsubsection{Batch size check ($N = 3$)}

@@

\subsubsection{Network architecture check ($N = 3$)}

@@

\subsubsection{Data leakage check ($N = 1$)}

@@

\subsubsection{Model performance check ($N = 33$)}

@@

\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}

@@

\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}

@@

\subsubsection{Spot check ($N = 5$)}

@@

\subsubsection{Model training check ($N = 4$)}

@@

\subsubsection{Missing value check ($N = 3$)}

Checking for missing values is a critical step in the preprocessing phase of an ML project because it significantly influences the quality and performance of the model. Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\citep{shome2022data}. Many ML algorithms demand complete numerical datasets to perform calculations such as matrix multiplication. Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely. These checks can be performed in code (\emph{P74, L36}) and using visualizations such as a heatmap (\emph{L12}).

@@

\subsubsection{Shape check ($N = 3$)}

@@

\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}

@@

Linear ML models achieve optimal performance when the target variable can be expressed as a linear combination of the input features. However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training. Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\citep{shome2022data}. Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships among features in the data. In addition to the code, the raw output of the cell is also shown since both case-studies produce a visualization.  In \emph{L6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}. \emph{L10} is a visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.

@@

\subsubsection{Type check ($N = 2$)}

@@

\subsubsection{Execution time check ($N = 1$)}
\begin{lstlisting}[caption={Print statement \emph{P66} used to check the total execution time of training an ML model with various hyper-parameter configurations.}, label={lst:P66}]
print('Total Run Time:')
\end{lstlisting}

While training multiple ML models for a single task is common, choosing the best model considers not only its performance but also its training speed. Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\citep{shome2022data}. Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyperparameter configurations using a grid-search.

@@

\subsubsection{Network architecture check ($N = 1$)}

@@

Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\citep{kery2018story,rule2018exploration,pimentel2019large-scale}. Once the code is finalized, it is typically transferred into a script that is run in a production environment. Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration. To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\citep{kery2018story,rule2018exploration}. In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\citep{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.

Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks. The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies. As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt. This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\citep{sculley2015hidden,amershi2019software,sambasivan2021everyone}.

Our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle. Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.

We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\citep{heer2010tour}. By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.

@@

Catching data errors are critical since ML models trained using dirty data will lead to poor and incorrect predictions. This is a cause for concern since these models may be deployed in production and their output may be used by other services which will then also perform poorly. The output of the ML models are often also used to capture new training data, thus leading to a feedback loop which over time will degrade the model performance~\citep{sculley2015hidden,breck2019data}.

\citet{breck2017ml} view ML training as analogous to compilation and recommend that training data be tested, similar to code in traditional software. They propose a rubric to assess the production readiness of an ML pipeline, which includes several recommendations for data testing. In subsequent work, \citet{breck2019data} introduce \emph{TensorFlow Data Validation}, an automated tool for data validation that works on principles borrowed from the database community. Data expectations are specified in a schema, and each new data batch is validated against this schema to ensure it meets the expected criteria or prompts schema updates if requirements change.

The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged. This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\citep{wang2020assessing}.

Results from RQ2 and RQ3 show that assertions, print statements and last cell statements are predominantly used in the upstream, data-intensive stages of the MLDL. For instance, this study identifies several assertions that perform data validation checks to ensure data meets specific criteria or constraints. Similarly, we find assertions, print statements, and last cell statements that conduct type, shape, and distribution checks to safeguard against data inconsistencies. These results show a missed opportunity in data validation tasks. We argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
